---
title: "RNA-seq experiment"
author: "Francesco Gastaldello"
date: "21 July 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup
Get *SpikeNorm* from GithHub private repository and preparare all necessary libraries:

```{r message=F}
require(SpikeNorm) #privately own package, every rights belong to the Data Analysis Group (http://www.compbio.dundee.ac.uk/dag.html)
require(tximport)
require(rhdf5)
require(edgeR)
```

Import data from SpikeNorm package and university cluster, preview the data and clean it accordingly.
```{r message=F}
data("exprsYeast")
data("factsYeast")

## Preview data
head(exprsYeast)
factsYeast

## strip out zero count genes in all samples
geneCounts <- exprsYeast[rowSums(exprsYeast)>0,]

#Import and prepare Salmon and Kallisto data
DIR<-'Z:/kallisto_test/out'
FILES <- file.path(DIR,c(paste0("output_Snf2-",1:6),paste0("output_WT",1:6)),"abundance.h5")
names(FILES) <- paste0("sample",1:12)
txk <- tximport(FILES,type="kallisto",txOut =T)
head(txk$counts)

DIR <- 'Z:/salmon_test/out'
FILES <- file.path(DIR,c(paste0("output_Snf2-",1:6),paste0("output_WT",1:6)),"quant.sf")
names(FILES) <- paste0("sample",1:12)
txs <- tximport(FILES,type="salmon",txOut =T,dropInfReps=TRUE) #no inferential replicates (bootstrap)
head(txs$counts)

STARgc <- read.csv('Z:/STAR_test/STARgenecounts.tsv',sep="\t",header=TRUE)
rownames(STARgc) <- STARgc$GeneName
STARgc$GeneName <- NULL
colnames(STARgc) <- c(paste0("Snf2_",1:6),paste0("WT",1:6))
```

Prepare DGE objects for all the tools used:

```{r}
d <- DGEList(geneCounts,group=factsYeast[names(exprsYeast), "condition"])
dgList_kallisto <- DGEList(counts=txk$counts,genes = rownames(txk$counts))
dgList_salmon <- DGEList(counts=txs$counts,genes = rownames(txs$counts))
dgList_STAR <- DGEList(counts = STARgc,genes = rownames(STARgc))

#Preparation and estimation of general linear model for the two datasets
sampleType <- c(rep("SNF2",6),rep("WT",6))
```

## Spike-ins part

Isolate spike-ins from EdgeR/Salmon/Kallisto/STAR datasets:

```{r}
colnames(d) <- paste0("sample",c(7:12,1:6))
d$counts <- d$counts[,c(7:12,1:6)]
spikes_TH <- cpm(d)[grep("^ERCC",rownames(d$counts)),]
spikes_kallisto <- cpm(dgList_kallisto)[grep("^ERCC",rownames(dgList_kallisto$counts)),]
spikes_salmon <- cpm(dgList_salmon)[grep("^ERCC",rownames(dgList_salmon$counts)),]
spikes_STAR <- cpm(dgList_STAR)[grep("^ERCC",rownames(dgList_STAR$counts)),]
```

Plot distribution of log10 count:

```{r warning=F, fig.height=14, fig.width=16}
par(mfrow=c(2,2))
boxplot(log10(spikes_TH),main="TopHat")
boxplot(log10(spikes_kallisto),main="kallisto")
boxplot(log10(spikes_salmon),main="salmon")
boxplot(log10(spikes_STAR),main="STAR")
```
```{r echo=F}
par(mfrow=c(1,1))
```

Compare medians by samples and by datasets:

```{r}
se <- function(x) sqrt(var(x)/length(x))
avgs_TH <- apply(spikes_TH,2,mean)
se_TH <- apply(spikes_TH,2,se)
med_TH <- apply(spikes_TH,2,median)
avgs_kallisto <- apply(spikes_kallisto,2,mean)
se_kallisto <- apply(spikes_kallisto,2,se)
med_kallisto <- apply(spikes_kallisto,2,median)
avgs_salmon <- apply(spikes_salmon,2,mean)
se_salmon <- apply(spikes_salmon,2,se)
med_salmon <- apply(spikes_salmon,2,median)
avgs_STAR <- apply(spikes_STAR,2,mean)
se_STAR <- apply(spikes_STAR,2,se)
med_STAR <- apply(spikes_STAR,2,median)

plot(med_TH,type="l",ylim=c(.5,5.5),xlab="Sample",ylab="Median")
arrows(1:ncol(spikes_TH),avgs_TH-se_TH,1:ncol(spikes_TH),avgs_TH+se_TH,length = .05,angle=90,code=3)
lines(med_kallisto,ylim=c(.5,5.5),col="red")
arrows(med_kallisto,avgs_kallisto-se_kallisto,med_kallisto,avgs_kallisto+se_kallisto,length = .05,angle=90,code=3)
lines(med_salmon,ylim=c(.5,5.5),col="green")
arrows(med_salmon,avgs_salmon-se_salmon,med_salmon,avgs_salmon+se_salmon,length = .05,angle=90,code=3)
lines(med_STAR,ylim=c(.5,5.5),col="purple")
arrows(med_STAR,avgs_STAR-se_STAR,med_STAR,avgs_STAR+se_STAR,length = .05,angle=90,code=3)
legend("topright",legend=c("TopHat","Kallisto","Salmon","STAR"),lty=c(1,1,1,1),lwd=c(2.5,2.5,2.5,2.5),col=c("black","red","green","purple"))
```

## Samples part

Samples isolation in TH/Salmon/Kallisto/STAR

```{r warning=F, fig.height=14, fig.width=16}
samples_TH <- cpm(d)[!grepl("^ERCC",rownames(d$counts)),]
samples_kallisto <- cpm(dgList_kallisto)[!grepl("^ERCC",rownames(dgList_kallisto$counts)),]
samples_salmon <- cpm(dgList_salmon)[!grepl("^ERCC",rownames(dgList_salmon$counts)),]
samples_STAR <- cpm(dgList_STAR)[!grepl("^ERCC",rownames(dgList_STAR$counts)),]

#Plot distribution of log10 count
par(mfrow=c(2,2))
boxplot(log10(samples_TH[,c(1,3,5,2,4,6,7,9,11,8,10,12)]),main="TopHat")
boxplot(log10(samples_kallisto[,c(1,3,5,2,4,6,7,9,11,8,10,12)]),ylim=c(-2,5),main="kallisto")
boxplot(log10(samples_salmon[,c(1,3,5,2,4,6,7,9,11,8,10,12)]),ylim=c(-2,5),main="salmon")
boxplot(log10(samples_STAR[,c(1,3,5,2,4,6,7,9,11,8,10,12)]),ylim=c(-2,5),main="STAR")
```
```{r echo=F}
par(mfrow=c(1,1))
```

## Spikenorm analysis workflow

Prepare datasets with gene lengths:

```{r}
gene.length <- read.csv(file = "./gene_length.csv")
gene.length$X <- NULL
spike.length <- read.csv(file = "./ERCC_lengths.tsv",sep="\t")
lengths <- rbind(gene.length,spike.length)
rownames(lengths) <- lengths$id
lengths$id <- NULL
lengths <- as.matrix(lengths)
```

Let's start with the TopHat dataset: 

```{r message=F}
res <- augDGEList(cpm(geneCounts), sampleData = factsYeast[colnames(geneCounts),], geneSeqs = lengths[rownames(geneCounts),])
res <- calcSpikeNorm(res, treat = "condition")
res

par(mfrow=c(1,2))
plotMixFits(res)
plotMixFits(res, ntype='mix')
```
```{r echo=F}
par(mfrow=c(1,1))
```
Use the SpikeNorm normalisation factors for edgeR

```{r}
res <- setNormFactors(res, ntype='sampleID')

design = model.matrix(~sampleType)
y = estimateGLMCommonDisp(res, design)
y = estimateGLMTrendedDisp(y, design)
y = estimateGLMTagwiseDisp(y, design)
fit <- glmFit(y,design)
lrt <- glmLRT(fit)
topTags(lrt)
```
Let's do Kallisto datasets:

```{r message=F}
colnames(dgList_kallisto$counts)<-c(paste0("WT_",1:6),paste0("Snf2_",1:6))
resK <- augDGEList(as.data.frame(dgList_kallisto$counts), sampleData = factsYeast[colnames(as.data.frame(dgList_kallisto$counts)),], geneSeqs = lengths[rownames(as.data.frame(dgList_kallisto$counts)),])
resK <- calcSpikeNorm(resK, treat = "condition")
resK

par(mfrow=c(1,2))
plotMixFits(resK)
plotMixFits(resK, ntype='mix')
resK <- setNormFactors(resK, ntype='sampleID')
```
```{r echo=F}
par(mfrow=c(1,1))
```
```{r}
sampleType<-c(rep("WT",6),rep("SNF2",6))
design = model.matrix(~sampleType)
yK = estimateGLMCommonDisp(resK, design)
yK = estimateGLMTrendedDisp(yK, design)
yK = estimateGLMTagwiseDisp(yK, design)
fitK <- glmFit(yK,design)
lrtK <- glmLRT(fitK)
topTags(lrtK)
```

Let's do Salmon dataset
```{r message=F}
colnames(dgList_salmon$counts)<-c(paste0("WT_",1:6),paste0("Snf2_",1:6))
resS <- augDGEList(as.data.frame(dgList_salmon$counts), sampleData = factsYeast[colnames(as.data.frame(dgList_salmon$counts)),], geneSeqs = lengths[rownames(as.data.frame(dgList_salmon$counts)),])
resS <- calcSpikeNorm(resS, treat = "condition")
resS

par(mfrow=c(1,2))
plotMixFits(resS)
plotMixFits(resS, ntype='mix')
```
```{r echo=F}
par(mfrow=c(1,1))
```
```{r}
resS <- setNormFactors(resS, ntype='sampleID')

sampleType<-c(rep("WT",6),rep("SNF2",6))
design = model.matrix(~sampleType)
yS = estimateGLMCommonDisp(resS, design)
yS = estimateGLMTrendedDisp(yS, design)
yS = estimateGLMTagwiseDisp(yS, design)
fitS <- glmFit(yS,design)
lrtS <- glmLRT(fitS)
topTags(lrtS)
```

Let's do STAR dataset

```{r message=F}
colnames(dgList_STAR$counts)<-c(paste0("Snf2_",1:6),paste0("WT_",1:6))
resST <- augDGEList(as.data.frame(dgList_STAR$counts), sampleData = factsYeast[colnames(as.data.frame(dgList_STAR$counts)),], geneSeqs = lengths[rownames(as.data.frame(dgList_STAR$counts)),])
resST <- calcSpikeNorm(resST, treat = "condition")
resST

par(mfrow=c(1,2))
plotMixFits(resST)
plotMixFits(resST, ntype='mix')
```
```{r echo=F}
par(mfrow=c(1,1))
```
```{r}
resST <- setNormFactors(resST, ntype='sampleID')

sampleType<-c(rep("WT",6),rep("SNF2",6))
design = model.matrix(~sampleType)
yST = estimateGLMCommonDisp(resST, design)
yST = estimateGLMTrendedDisp(yST, design)
yST = estimateGLMTagwiseDisp(yST, design)
fitST <- glmFit(yST,design)
lrtST <- glmLRT(fitST)
topTags(lrtST)
```
## Further analysis

P-values vs Fold Change
```{r}
par(mfrow=c(2,2))
plot(lrt$table$PValue,lrt$table$logFC,xlab="P-Values",ylab="Fold Change",main="TopHat")
plot(lrtK$table$PValue,lrtK$table$logFC,xlab="P-Values",ylab="Fold Change",main="Kallisto")
plot(lrtS$table$PValue,lrtS$table$logFC,xlab="P-Values",ylab="Fold Change",main="Salmon")
plot(lrtST$table$PValue,lrtST$table$logFC,xlab="P-Values",ylab="Fold Change",main="STAR")
```
```{r echo=F}
par(mfrow=c(1,1))
```
